When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-05-14 doc/paper/vpd_et_paper.tex
some tweaking of figure and positions for arkiv submission
arkiv submission

\input{def}
\usepackage{makecell}
\usepackage{multirow} %AM for tables
\usepackage{booktabs}
\usepackage{authblk}
\usepackage{textcomp}

\bibliographystyle{agufull08}
\RequirePackage{natbib}
\RequirePackage{graphicx}

\def\acknowledgments{\vskip12pt\noindent{\bf
Acknowledgments\vrule depth 6pt
width0pt\relax}\\*\noindent\ignorespaces}

% \drafttrue

% \journalname{Agricultural and Forest Meteorology: When does VPD drive
%   or reduce ET?}

\newcommand{\appropto}{\mathrel{\vcenter{
      \offinterlineskip\halign{\hfil$##$\cr
        \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}

@@

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the
FLUXNET2015 database (data available at \sloppy
https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy),
including all sites with at least four years of data and observations
of the variables described in the methods section (Section
\ref{methods}). Seventy-three sites met these requirements, and were
grouped into nine plant functional types (PFT) according to the
International Geosphere-Biosphere Programme vegetation classification
scheme \citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous
broadleaf forest (DBF), evergreen broadleaf forest (EBF), evergreen
needleleaf forest (ENF), mixed forest (MF), closed shrub (CSH),
savannah (SAV), and woody savannah (WSA) (site locations in Figure
\ref{map_fig}).

\begin{figure}
  \centering \includegraphics{./map.pdf}
  \caption{Plant functional type and location of FLUXNET2015 sites
    used in this analysis.}
  \label{map_fig}
\end{figure}

@@

    \hline
    \input{pft_params}
    \hline
  \end{tabular}
\end{table}

@@

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published here. This makes Equation \ref{et} a useful alternative to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation \ref{et} better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), contrary
to typical drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016, Lemordant_2016, Lemordant_2018}. Lastly, all
code and data used in this analysis, including those used to generate
the figures and tables, are publicly available at
https://github.com/massma/climate\_et.

@@

\begin{figure}
  \centering \includegraphics{./joint_vpd_sigma.pdf}
  \caption{The joint distribution of $VPD$ and $\sigma$, with outliers
    removed (defined as lowest and highest 5\% of $\sigma$). $\sigma$
    exhibits a weak dependence on $VPD$, and $\sigma$ is $O(1)$ for
    the bulk of the observations.}
  \label{joint_vpd_sigma}
\end{figure}

@@

Biases between the theory and observations are similar for ENF, WSA,
CRO and SAV. At low VPD the theory and the observations match
well. However, at high VPD (upper 10-20\% of observations) the theory
is biased slightly towards positive response as compared to the
observations (Figure \ref{test_sign}). For ENF this slight bias could
be explained by a negative bias in $g_1$. However, for WSA, CRO, and
SAV the observations in the highest VPD bins exhibit a downturn
towards more negative response, which cannot be captured by the
functional form of the sign term (Figure
\ref{idealized_sign}). Therefore, to explain the observations one of
the variables in the sign term must change at extreme VPD (upper
10-20\%). The most likely candidate is $uWUE$, which we have assumed
constant to meet our goal of developing intuition for leading order
behavior, but might be expected to decrease at extremely low SWC. This extremely low SWC should be correlated with high VPD in very dry environments
through sensible heating increases \citep{Gentine_2016}. The theory
for EBF exhibits similar limitations as for ENF, WSA, and SAV, except
a larger portion of observations are biased negative as compared to
the theory ($\approx$ 35\%, Figure \ref{test_sign}). However, in
general, and specifically for non-extreme VPD (VPD < $\approx$70-90th
percentiles), the theory matches the observations for the tree and
savannah plant types well.

@@

While the above discussion shows that our theory has some limitations
when applied to some PFTs, especially for grasslands, it does well for
DBF, MF and CSH PFTs, and captures the response at non-extreme VPD
(VPD < $\approx$ 70th-90th percentiles) for
ENF, EBF, CRO, WSA, and SAV. In general, the leading order behavior
observed in the data is captured by the theory. Departures between our
theory and observations, specifically at high VPD, could be explained
and conceptualized with shifts in $g_1$ and/or $uWUE$ due to
site-specific plant type variability (e.g. more arid-adapted ecosystems
at more arid sites), or temporal variability for some environmental
conditions (e.g. decreases in $uWUE$ at extremely low soil
moisture). To focus on general behavior and develop intuition for
PFT-scale response, we have ignored these sources of variability in
the present analysis. However, any site-specific or temporal plant
functional variability that can be conceptualized with shifts to $g_1$
and $uWUE$ can be analyzed with our framework. This opens a door to future analyses in which plant behavior in anomalous conditions
can be explained and analyzed using Equation \ref{et}.

@@

To summarize, our theory is limited by its inability to account for
soil water impacts on land surface response, and feedbacks between SWC
and VPD. Fortunately, for most PFTs SWC's effect on ET is relatively
rare (<30\% of observations) and does not manifest itself in the
majority of observations and bulk statistics. However for GRA, SWC decreases water use efficiency for the majority of
the observations. Soil moisture effects explain the deficiencies of
our theory in Section \ref{testing}, particularly for GRA. By
conceptualizing SWC effects as a change in $uWUE$ (and/or $g_1$), it
will be possible for future analysis to explore the importance of soil
moisture on plant response to VPD, and feedbacks between plant
function, SWC, and VPD.

@@

\begin{figure}
  \centering
  \centerline{\includegraphics[width=1\textwidth]{./concave.pdf}}
  \caption{ Solutions corresponding to inflection points between
    concave up and concave down ET-VPD curves (Equation \ref{curves})
    for three specific scenarios. Solutions are defined in terms of a
    non-dimensional VPD ($VPD^m/g_*$), but to aide physical
    interpretation the horizontal axis is additionally provided in terms of dimensionalized VPD assuming $m=1/2$ and
    $g_*=110\; Pa^{1/2}$ (average of all PFT $g_1$). The vertical axis
    has a different interpretation depending on the solution
    curve. For the blue line ($m$ varying), it corresponds to $m$,
    for the orange line ($n$ varying) it corresponds to $n$, and for
    the green line it corresponds to the value of both $n$ and $m$
    ($n=m$). Regions of the parameter space that correspond to
    concave up and concave down results are labeled.}
  \label{concave}
\end{figure}

@@

\acknowledgments This work used eddy covariance data acquired and
shared by the FLUXNET community, including these networks: AmeriFlux,
AfriFlux, AsiaFlknux, CarboAfrica, CarboEuropeIP, CarboItaly, CarboMont,
ChinaFlux, Fluxnet-Canada, GreenGrass, ICOS, KoFlux, LBA, NECC,
OzFlux-TERN, TCOS-Siberia, and USCCC. The ERA-Interim reanalysis data
are provided by ECMWF and processed by LSCE. The FLUXNET eddy
covariance data processing and harmonization was carried out by the
European Fluxes Database Cluster, AmeriFlux Management Project, and
Fluxdata project of FLUXNET, with the support of CDIAC and ICOS
Ecosystem Thematic Center, and the OzFlux, ChinaFlux and AsiaFlux
offices. This material is based upon work supported by the National
Science Foundation Graduate Research Fellowship under Grant No. DGE
16-44869. All code and data used in this analysis, including those
used to generate the figures and tables, are publicly available at
https://github.com/massma/climate\_et.

>>>>

\documentclass[12pt]{article}
\input{def.tex}

@@

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the
FLUXNET2015 database (data available at \sloppy
\url{https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/} \sloppy),
including all sites with at least four years of data and observations
of the variables described in the methods section (Section
\ref{methods}). Seventy-three sites met these requirements, and were
grouped into nine plant functional types (PFT) according to the
International Geosphere-Biosphere Programme vegetation classification
scheme \citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous
broadleaf forest (DBF), evergreen broadleaf forest (EBF), evergreen
needleleaf forest (ENF), mixed forest (MF), closed shrub (CSH),
savannah (SAV), and woody savannah (WSA) (site locations in Figure
\ref{map_fig}).

\begin{figure}[h]
  \centering \includegraphics[trim={0 3cm 0 3cm}, clip]{./map.pdf}
  \caption{Plant functional type and location of FLUXNET2015 sites
    used in this analysis.}
  \label{map_fig}
\end{figure}

@@

    \hline
    \input{pft_params.tex}
    \hline
  \end{tabular}
\end{table}

@@

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published here. This makes Equation \ref{et} a useful alternative to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation \ref{et} better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), contrary
to typical drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016, Lemordant_2016, Lemordant_2018}. Lastly, all
code and data used in this analysis, including those used to generate
the figures and tables, are publicly available at
\url{https://github.com/massma/climate\_et}.
%\hyperref[https://github.com/massma/climate\_et]{https://github.com/massma/climate_et}.

@@

\begin{figure}[h]
  \centering\includegraphics[width=0.75\textwidth]{./joint_vpd_sigma.pdf}
  \caption{The joint distribution of $VPD$ and $\sigma$, with outliers
    removed (defined as lowest and highest 5\% of $\sigma$). $\sigma$
    exhibits a weak dependence on $VPD$, and $\sigma$ is $O(1)$ for
    the bulk of the observations.}
  \label{joint_vpd_sigma}
\end{figure}

@@

Biases between the theory and observations are similar for ENF, WSA,
CRO and SAV. At low VPD the theory and the observations match
well. However, at high VPD (upper 10-20\% of observations) the theory
is biased slightly towards positive response as compared to the
observations (Figure \ref{test_sign}). For ENF this slight bias could
be explained by a negative bias in $g_1$. However, for WSA, CRO, and
SAV the observations in the highest VPD bins exhibit a downturn
towards more negative response, which cannot be captured by the
functional form of the sign term (Figure
\ref{idealized_sign}). Therefore, to explain the observations one of
the variables in the sign term must change at extreme VPD (upper
10-20\%). The most likely candidate is $uWUE$, which we have assumed
constant to meet our goal of developing intuition for leading order
behavior, but might be expected to decrease at extremely low SWC. This extremely low SWC should be correlated with high VPD in very dry environments
through sensible heating increases \citep{Gentine_2016}. The theory
for EBF exhibits similar limitations as for ENF, WSA, and SAV, except
a larger portion of observations are biased negative as compared to
the theory ($\approx$ 35\%, Figure \ref{test_sign}). However, in
general, and specifically for non-extreme VPD (VPD  $< \approx$70-90th
percentiles), the theory matches the observations for the tree and
savannah plant types well.

@@

While the above discussion shows that our theory has some limitations
when applied to some PFTs, especially for grasslands, it does well for
DBF, MF and CSH PFTs, and captures the response at non-extreme VPD
(VPD $< \approx$ 70th-90th percentiles) for
ENF, EBF, CRO, WSA, and SAV. In general, the leading order behavior
observed in the data is captured by the theory. Departures between our
theory and observations, specifically at high VPD, could be explained
and conceptualized with shifts in $g_1$ and/or $uWUE$ due to
site-specific plant type variability (e.g. more arid-adapted ecosystems
at more arid sites), or temporal variability for some environmental
conditions (e.g. decreases in $uWUE$ at extremely low soil
moisture). To focus on general behavior and develop intuition for
PFT-scale response, we have ignored these sources of variability in
the present analysis. However, any site-specific or temporal plant
functional variability that can be conceptualized with shifts to $g_1$
and $uWUE$ can be analyzed with our framework. This opens a door to future analyses in which plant behavior in anomalous conditions
can be explained and analyzed using Equation \ref{et}.

@@

To summarize, our theory is limited by its inability to account for
soil water impacts on land surface response, and feedbacks between SWC
and VPD. Fortunately, for most PFTs SWC's effect on ET is relatively
rare ($<$30\% of observations) and does not manifest itself in the
majority of observations and bulk statistics. However for GRA, SWC decreases water use efficiency for the majority of
the observations. Soil moisture effects explain the deficiencies of
our theory in Section \ref{testing}, particularly for GRA. By
conceptualizing SWC effects as a change in $uWUE$ (and/or $g_1$), it
will be possible for future analysis to explore the importance of soil
moisture on plant response to VPD, and feedbacks between plant
function, SWC, and VPD.

@@

\begin{figure}
  \centering
  \centerline{\includegraphics[width=0.75\textwidth]{./concave.pdf}}
  \caption{ Solutions corresponding to inflection points between
    concave up and concave down ET-VPD curves (Equation \ref{curves})
    for three specific scenarios. Solutions are defined in terms of a
    non-dimensional VPD ($VPD^m/g_*$), but to aide physical
    interpretation the horizontal axis is additionally provided in terms of dimensionalized VPD assuming $m=1/2$ and
    $g_*=110\; Pa^{1/2}$ (average of all PFT $g_1$). The vertical axis
    has a different interpretation depending on the solution
    curve. For the blue line ($m$ varying), it corresponds to $m$,
    for the orange line ($n$ varying) it corresponds to $n$, and for
    the green line it corresponds to the value of both $n$ and $m$
    ($n=m$). Regions of the parameter space that correspond to
    concave up and concave down results are labeled.}
  \label{concave}
\end{figure}

@@

\acknowledgments This work used eddy covariance data acquired and
shared by the FLUXNET community, including these networks: AmeriFlux,
AfriFlux, AsiaFlknux, CarboAfrica, CarboEuropeIP, CarboItaly, CarboMont,
ChinaFlux, Fluxnet-Canada, GreenGrass, ICOS, KoFlux, LBA, NECC,
OzFlux-TERN, TCOS-Siberia, and USCCC. The ERA-Interim reanalysis data
are provided by ECMWF and processed by LSCE. The FLUXNET eddy
covariance data processing and harmonization was carried out by the
European Fluxes Database Cluster, AmeriFlux Management Project, and
Fluxdata project of FLUXNET, with the support of CDIAC and ICOS
Ecosystem Thematic Center, and the OzFlux, ChinaFlux and AsiaFlux
offices. This material is based upon work supported by the National
Science Foundation Graduate Research Fellowship under Grant No. DGE
16-44869. All code and data used in this analysis, including those
used to generate the figures and tables, are publicly available at
\url{https://github.com/massma/climate\_et}.
%\hyperref[https://github.com/massma/climate\_et]{https://github.com/massma/climate_et}.
